\documentclass[pra,11pt]{revtex4}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color}
\usepackage{mathrsfs}

\def\ket#1{\left|#1\right\rangle}
\def\bra#1{\left\langle#1\right|}
\def\braket#1{\left\langle#1\right\rangle}

\setlength{\parindent}{0pt}

\renewcommand{\baselinestretch}{1.0}
\setlength{\parskip}{0.07in}

\begin{document}

\section{Quantum mechanics of multi-particle systems}

So far, we have studied quantum mechanical systems that consist of
single particles.  The next important step is to look at systems of
more than one particle.  As we shall see, the postulates of quantum
mechanics have interesting implications which only show up in
multi-particle systems, such as the phenomenon of \textbf{quantum
  entanglement}.

In the following discussion, you may assume that we are dealing with
systems of ``distinguishable'' particles, such as a two-particle
system of an electron and a proton.  (There is a separate set of
complications whenever we deal with ``identical'' or
``indistinguishable'' particles, such as a system of two electrons.
We will spend the next chapter dealing with those issues.  If you're
not sure what this means, don't worry; just read on.)

Suppose we have two distinguishable particles, labeled $A$ and $B$.
If each individual particle is treated as a quantum system, then
according to the usual postulates of quantum mechanics, its state is
described by a vector in some complex Hilbert space.  Let
$\mathscr{H}_A$ and $\mathscr{H}_B$ denote the respective
single-particle Hilbert spaces.  When the two particles are considered
as a single system, the combined Hilbert space is denoted by
$$\mathscr{H} = \mathscr{H}_A\otimes \mathscr{H}_B.$$
The symbol $\otimes$ refers to a mathematical operation known as the
\textbf{tensor product}, which is a way of combining two Hilbert
spaces to form a larger Hilbert space.

The mathematical meaning of the tensor product is as follows.  Suppose
$\mathscr{H}_A$ can be spanned by an orthonormal basis
$\{|\mu_1\rangle, |\mu_2\rangle, |\mu_3\rangle, \dots\}$, and
$\mathscr{H}_B$ can be spanned by $\{|\nu_1\rangle, |\nu_2\rangle,
|\nu_3\rangle, \dots\}$.  Then the tensor product space $\mathscr{H}_A
\otimes \mathscr{H}_B$ can be defined in terms of the basis set
$$\Big\{\;\,|\mu_i\rangle\otimes|\nu_j\rangle \;\;  \Big| \;\; \textrm{all}\;|\mu_i\rangle,\; |\nu_j\rangle \;\,\Big\}.$$
Intuitively, this means that if you know particle $A$ is in state
$|\mu_i\rangle$, and particle $B$ is in state $|\nu_j\rangle$, that
specifies a state of the combined system.  A complete basis set for
the combined system can be constructed in this way.  Note that if
$\mathscr{H}_A$ has dimension $d_A$ and $\mathscr{H}_B$ has dimension
$d_B$, then $\mathscr{H}_A \otimes \mathscr{H}_B$ has dimension $d_A
d_B$.

Any two-particle state can then be written as
$$|\psi\rangle = \sum_{i} \sum_{j} \, c_{ij}\; |\mu_i\rangle \otimes |\nu_j\rangle.$$
The inner product for tensor product basis states is defined in terms
of the inner product of the individual spaces:
$$\Big(|\mu_i\rangle \otimes |\nu_j\rangle\;,\; |\mu_p\rangle \otimes |\nu_q\rangle \Big) \;\equiv\; \big(\langle\mu_i| \otimes \langle\nu_j| \big) \big(|\mu_p\rangle \otimes |\nu_q\rangle\big) \;\equiv\; \langle\mu_i|\mu_p\rangle \, \langle\nu_j|\nu_q\rangle.$$
In other words, to multiply a tensor product bra with a tensor product
ket, (i) do the bra-ket product for the first slot, (ii) do the
bra-ket product for the second slot, and (iii) multiply the results
together.  You may check that this satisfies the requirements for an
inner product in linear algebra (assuming that the inner products for
the single-particle states are well-defined).

As an example, suppose $\mathscr{H}_A$ and $\mathscr{H}_B$ are both
two-dimensional Hilbert spaces describing spin-$1/2$ degrees
of freedom, such that each can be spanned by an orthonormal basis
$\{\,|\!+\!z\rangle, \,|\!-\!z\rangle \, \}$, representing ``spin-up''
and ``spin-down'' states.  Then the tensor product space $\mathscr{H}$
is spanned by
$$\Big\{\;|\!+\!z\rangle\otimes|\!+\!z\rangle\,,\; |\!+\!z\rangle\otimes|\!-\!z\rangle\,,\; |\!-z\!\rangle\otimes|\!+\!z\rangle\,,\; |\!-\!z\rangle\otimes|\!-\!z\rangle \;\Big\}.$$
The dimension of $\mathscr{H}$ is $2 \times 2 = 4$.  

At this point, we make an important observation.  If we specify that
particle $A$ is in state $|\mu_i\rangle$ and particle $B$ is in state
$|\nu_j\rangle$, that specifies a state of the combined system; but
the reverse is not true!  \textit{Some states of the combined system
  cannot be expressed in terms of definite states of the individual
  particles.}  For example, take the previous example of two spaces
describing spin-$1/2$ degrees of freedom, and consider
$$|\psi\rangle = \frac{1}{\sqrt{2}} \Big(|\!+\!z\rangle\otimes|\!-\!z\rangle \,-\, |\!-z\!\rangle\otimes|\!+\!z\rangle\Big).$$

This two-particle state is constructed from two of the basis states of
the tensor product space,
$|\!+\!z\rangle\otimes|\!-\!z\rangle$ and
$|\!-\!z\rangle\otimes|\!+\!z\rangle$; you can check that
the factor of $1/\sqrt{2}$ correctly normalizes the state so that
$\langle\psi|\psi\rangle = 1$.  Evidently, neither particle A nor
particle B has a definite $|\!+\!z\rangle$ or
$|\!-\!z\rangle$ state.  Furthermore, we shall show rigorously,
in a later section, that there's \textit{no} choice of basis in which this
particular quantum state $|\psi\rangle$ can be expressed with
definite quantum states for each particle.  In other words,
$$|\psi\rangle \ne |\psi_A\rangle\otimes|\psi_B\rangle \;\;\;\textrm{for}\;\textrm{any}\;\; |\psi_A\rangle \in \mathscr{H}_A, \;|\psi_B\rangle \in \mathscr{H}_B.$$
In such a situation, the two particles are said to be
\textbf{entangled}.  This means that they cannot be described by
individual single-particle quantum states, but are intrinsically
``mixed together''.

For systems of more than two particles, the multi-particle quantum
states are defined using multiple tensor products.  Suppose we have
$N$ particles, with single-particle Hilbert spaces $\mathscr{H}_1,
\mathscr{H}_2, \dots, \mathscr{H}_N$, of dimensionality $d_1, \dots,
d_N$.  Then the multi-particle Hilbert space is
$$\mathscr{H} = \mathscr{H}_1 \otimes \mathscr{H}_2 \otimes \cdots
\otimes \mathscr{H}_N,$$
which has dimensionality $d = d_1 d_2\cdots d_N$.  It is interesting
to note that the dimensionality of the multi-particle Hilbert space
scales \textit{exponentially} with the number of particles.  For
instance, if each single-particle Hilbert space is two-dimensional, a
$20$-particle system will have a Hilbert space of dimensionality
$2^{20} =1\,048\,576$.  This means that multi-particle quantum systems
are able to ``carry'' huge amounts of information, compared to
classical systems.  That is one of the reasons people are excited by
the prospect of quantum computing.

\section{Partial measurements}
\label{sec:partialmeasurements}

Let us recall how measurements work in single-particle quantum theory.
Suppose a particle is in a quantum state $|\psi\rangle$ and we perform
some measurement $\hat{Q}$.  To figure out the measurement outcome, we
first write $|\psi\rangle$ in the eigenstate basis of $\hat{Q}$:
$$|\psi\rangle = \sum_n \psi_n\, |q_n\rangle, \;\;\mathrm{where}\;\;\hat{Q}|q_n\rangle = q_n |q_n\rangle \;\,\textrm{and}\;\, \psi_n = \langle q_n|\psi\rangle.$$
For simplicity, let's suppose the eigenvalues $\{q_n\}$ are all
non-degenerate.  Then the probability of obtaining the measurement
outcome $q_n$ is $|\psi_n|^2$, the absolute square of the coefficient
of $|q_n\rangle$ in the above formula.  After the measurement, the
system ``collapses'' into state $|q_n\rangle$, which can be regarded as
the projection operation
$$|\psi\rangle \longrightarrow \frac{1}{\sqrt{\mathcal{N}}}\; |q_n\rangle\langle q_n|\psi\rangle.$$
The prefactor $\mathcal{N} = |\psi_n|^2$ simply ensures that the new
state remains normalized; the overall phase has no physical
significance.

For multi-particle systems, the measurement postulates of quantum
theory are faced with a new complication: how do we describe a
measurement performed on just one particle?

To figure it out, let's again consider a system of two particles, with
Hilbert spaces $\mathscr{H}_A$ and $\mathscr{H}_B$, and the
two-particle Hilbert space $\mathscr{H} = \mathscr{H}_A \otimes
\mathscr{H}_B$.  Suppose we perform a measurement on particle $A$,
corresponding to a Hermitian operator $\hat{\mu}$ that acts upon
$\mathscr{H}_A$ and has eigenvectors $\{|\mu_1\rangle,
|\mu_2\rangle,\dots\}$.  We can write any two-particle state
$|\psi\rangle$ using the eigenvectors of $\hat{Q}$ as a basis set for
$\mathscr{H}_A$, and an arbitrary basis set $\{|\nu_1\rangle,
|\nu_2\rangle,\dots\}$ for $\mathscr{H}_B$:
$$\begin{aligned}|\psi\rangle &= \sum_{i}\sum_{j} \psi_{ij}\, |\mu_i\rangle\otimes |\nu_j\rangle \\&= \sum_i |\mu_i\rangle\otimes |\varphi_i\rangle, \;\;\mathrm{where}\;\;|\varphi_i\rangle\equiv \sum_j \psi_{ij}\,|\nu_j\rangle \;\in\; \mathscr{H}_B.\end{aligned}$$
Note that unlike the single-particle case, the ``coefficient'' of
$|n\rangle$ in the basis expansion is not merely a complex number, but
a vector in the space $\mathscr{H}_B$.  We can proceed by
analogy.  The probability of obtaining the measurement outcome $\mu_i$
should be the ``absolute square'' of this ``coefficient'':
$$\langle \varphi_i|\varphi_i\rangle = \sum_j |\psi_{ij}|^2.$$
After we obtain the measurement result $\mu_i$, the state collapses
via to the projection
$$|\psi\rangle \longrightarrow \frac{1}{\sqrt{\mathcal{N}}}\; \Big(|\mu_i\rangle\langle \mu_i| \otimes \hat{I}\Big) |\psi\rangle = \frac{1}{\sqrt{\mathcal{N}}}\; |\mu_i\rangle\otimes |\varphi_i\rangle.$$
Note that the projection operator $|\mu_i\rangle\langle \mu_i|$ acts
only upon the $\mathscr{H}_A$ part of the two-particle space.  The
factor $\mathcal{N}$ is defined so that the resulting state is
normalized to unity.

Let's work through an explicit example, involving the
previously-discussed system of two spin-$1/2$ particles, and the
two-particle entangled state
$$|\psi\rangle = \frac{1}{\sqrt{2}} \Big(|\!+\!z\rangle\otimes|\!-\!z\rangle \,-\, |\!-\!z\rangle\otimes|\!+\!z\rangle\Big).$$
Here, $|\!+\!z\,\rangle$ and $|\!-\!z\,\rangle$ denote eigenstates
of the operator $\hat{S}_z$, with the eigenvalues $+\hbar/2$ and $-\hbar/2$
respectively.  Let us perform an $S_z$ measurement on particle A.  The
results are as follows:
\begin{itemize}
\item A measurement outcome of $+\hbar/2$ has probability $P_+ = \langle
  n'|n'\rangle$, where $|n'\rangle =
  (1/\sqrt{2})\,|\!-\!z\rangle$.  Hence, $P_+ = 1/2$.  After the
  measurement, the state collapses to $|\!+\!z\rangle
  \otimes|\!-\!z\rangle$.

\item A measurement outcome of $-\hbar/2$ has probability $P_- = \langle
  n'|n'\rangle$, where $|n'\rangle =
  (1/\sqrt{2})\,|\!+\!z\rangle$.  Hence, $P_- = 1/2$.  After the
  measurement, the state collapses to $|\!-\!z\rangle
  \otimes|\!+\!z\rangle$.
\end{itemize}
In other words, we obtain the two possible results $\pm \hbar/2$ with
equal probability; in either case, the two-particle state collapses so
that particle $A$ is in the spin eigenstate corresponding to our
measurement result, while particle $B$ has the opposite spin.  After
collapse, the two-particle state is no longer entangled.

\section{The Einstein-Podolsky-Rosen ``paradox''}

In 1935, Einstein, Podolsky, and Rosen (EPR) formulated a thought
experiment known as the \textbf{EPR paradox}, which highlighted the
exceptionally counter-intuitive features of entangled quantum states.
They argued that because of these features, quantum theory had to be
an incomplete theory of reality.  Subsequently, however, it was shown
that the EPR paradox is not an actual paradox; physical systems
\textit{really do} behave in the counter-intuitive way predicted by
quantum mechanics.

The EPR thought experiment begins with an entangled state, such as the
entangled spin-$1/2$ two-particle state described in the previous section:
$$|\psi\rangle = \frac{1}{\sqrt{2}} \Big(|\!+\!z\rangle\otimes|\!-\!z\rangle \,-\, |\!-\!z\rangle\otimes|\!+\!z\rangle\Big).$$
As previously discussed, we can measure $\hat{S}_z$ on particle $A$.
The system then collapses into a two-particle state with definite
spins: if we measured $+\hbar/2$, the collapsed state is
$|\!+\!z\rangle \otimes|\!-\!z\rangle$, and if we measured $-\hbar/2$,
the collapsed state is $|\!-\!z\rangle\otimes|\!+\!z\rangle$.

According to quantum theory, the state collapse happens
instantaneously, regardless of the distance between the particles.  We
could prepare the two-particle state in a laboratory and hang on to
particle $A$, while sending particle $B$ to the Betelgeuse star
system, 642 light years away.  In principle, this can be done
carefully enough to avoid disturbing the two-particle quantum state.
Once ready, we measure $\hat{S}_z$ on particle $A$, which induces an
instantaneous collapse of the two-particle state.  Immediately
afterwards, if our alien colleague at Betelgeuse measures $\hat{S}_z$
on particle $B$, xhe will obtain (with 100\% certainty) the opposite
of our Earth result.  Yet in the time between these two measurements,
no classical signal could have traveled between Earth and Betelgeuse,
not even at the speed of light.

\textcolor{red}{[Fig.]}

There are three noteworthy aspects of this phenomenon.

First, it dispels popular intuitive ``explanations'' of quantum state
collapse that involve experimental probes physically disturbing the
system being measured.  For instance, one story that's sometimes told
is that measuring a particle's position requires shining a light beam
on it, or disturbing it in some way; because of this disturbance, the
particle's momentum becomes uncertain.  The EPR paradox shows that
such stories don't capture the full weirdness of quantum state
collapse.  We can collapse the state of a particle at Betelgeuse by
doing a measurement on another particle, hundreds of light years away!

Second, the experimentalist has control over what kind of measurement
to perform.  So far, we have considered $\hat{S}_z$ measurements
performed on particle $A$.  But the Earth experimentalist can choose
to measure the spin of $A$ along any axis, say $\hat{S}_x$.  In the
basis of spin-up and spin-down states, the $\hat{S}_x$ operator has
the matrix representation
$$\hat{S}_x = \frac{\hbar}{2}\, \begin{pmatrix}0&1\\1&0\end{pmatrix}.$$
The eigenvalues and eigenvectors are
$$\begin{aligned}s_x = \;\;\frac{\hbar}{2},\; &\;\;\; |\!+\!x\rangle = \frac{1}{\sqrt{2}}\Big(|\!+\!z\rangle + |\!-\!z\rangle\Big) \\ s_x = -\frac{\hbar}{2}, &\;\;\; |\!-\!x\rangle = \frac{1}{\sqrt{2}}\Big(|\!+\!z\rangle - |\!-\!z\rangle\Big).\end{aligned}$$
Conversely, we can write the $\hat{S}_z$ eigenstates in the $\{|\!+\!x\rangle,|\!-\!x\rangle\}$ basis:
$$\begin{aligned}|\!+\!z\rangle &= \frac{1}{\sqrt{2}}\Big(|\!+\!x\rangle + |\!-\!x\rangle\Big) \\ |\!-\!z\rangle &= \frac{1}{\sqrt{2}}\Big(|\!+\!x\rangle - |\!-\!x\rangle\Big).\end{aligned}$$
This allows us to write the two-particle entangled state in the
$\hat{S}_x$ basis:
$$|\psi\rangle = \frac{1}{\sqrt{2}} \Big(|\!-\!x\rangle\otimes|\!+\!x\rangle \,-\, |\!+\!x\rangle\otimes|\!-\!x\rangle\Big).$$
The Earth measurement causes the particles to collapse into definite
spin states with opposite spins---but these are now spin states of
definite ${S}_x$, rather than definite ${S}_z$.

Third, the choice of measurement cannot be used for superluminal
communication.  The Earth experimentalist can choose whether to (i)
measure $\hat{S}_z$ or (ii) measure $\hat{S}_x$, and this choice has
an instantaneous effect on the quantum state of particle $B$.  If the
Betelgeuse experimentalist can find a way to distinguish between case
(i) and case (ii), even statistically, this could serve as a method of
instantaneous communication.  Yet this turns out to be impossible!
This is because quantum states themselves cannot be measured; only
observables can be measured.  Suppose the Earth measurement is
$\hat{S}_z$, which collapses $B$ to either $|\!+\!z\rangle$ or
$|\!-\!z\rangle$ (each with probability 0.5).  The Betelgeuse
experimentalist now has a choice of which measurement to perform.  If
xhe measures $\hat{S}_z$, the outcome is $+\hbar/2$ or $-\hbar/2$ with
equal probabilities.  Whereas for an $\hat{S}_x$ measurement, the
probabilities are:
$$\begin{aligned}P(S_x = +\hbar/2) &= \frac{1}{2}\, \Big|\langle\!+x|\!+\!z\rangle\Big|^2 \;+\;\, \frac{1}{2}\, \Big|\langle\!+x|\!-\!z\rangle\Big|^2 = \frac{1}{2}\\P(S_x = -\hbar/2) &= \frac{1}{2}\, \Big|\langle\!-x|\!+\!z\rangle\Big|^2 \;+\;\, \frac{1}{2}\, \Big|\langle\!-x|\!-\!z\rangle\Big|^2 = \frac{1}{2}.\end{aligned}$$
This analysis can be repeated for the case where the Earth measurement
is $\hat{S}_x$.  No matter what, we find that the outcomes of any spin
measurement at Betelgeuse always have 50/50 probability.  Thus, even
though the choice of Earth measurement instantaneously affects the
quantum state at Betelgeuse, no measurement at Betelgeuse can provide
any information about that choice.

Since quantum state collapse does not allow for superluminal
communication, it is consistent, \textit{in practice}, with the theory
of relativity.  However, state collapse is still
``\textbf{nonlocal}'', in the sense that certain ingredients of the
theory (quantum states) change faster than light can travel between
two points.  For this reason, EPR argued that quantum theory is
\textit{philosophically inconsistent} with relativity.  They
suggested, as an alternative, that quantum mechanics is an
approximation of some deeper theory, whose details are currently not
known to us, but which is both deterministic and local.

Such a ``\textbf{hidden variables theory}'' might give rise to the
appearance of quantum state collapse in the following way.  Suppose
each particle has a definite but ``hidden'' value of $S_z$, either
$S_z = +\hbar/2$ or $S_z = -\hbar/2$; for conciseness, we denote these
as $+$ or $-$.  We then hypothesize that the two-particle quantum state
$$|\psi\rangle = \frac{1}{\sqrt{2}} \Big(|\!+\!z\rangle\otimes|\!-\!z\rangle \,-\, |\!-\!z\rangle\otimes|\!+\!z\rangle\Big)$$
is not an actual description of reality.  Instead, it corresponds to a
\textit{statistical} distribution of ``hidden variable'' states, which
we denote by $[+;-]$ (i.e., $S_z = +\hbar/2$ for particle $A$ and $S_z
= -\hbar/2$ for particle $B$), and $[-;+]$ (the other way around).
When the Earth experimentalist measures $S_z$, we uncover the value of
the hidden variables: a $+z$ result implies $[+;-]$, and a $-z$ result
implies $[-;+]$.  When the Betelgeuse experimentalist subsequently
measures $S_z$, the result obtained is the opposite of the Earth
result---simply because those were the values of the hidden variables
all along.  No physical influence actually travels from Earth to
Betelguese between the two experiments.

Of course, there is a huge amount of detail missing about the actual
contents of the ``hidden variables'' theory.  In particular, such a
theory would have to be consistent, in all other respects, with the
many successful predictions of quantum mechanics.  This seems
difficult, but one could imagine that with enough hard work, it might
be possible to fill in the missing details.

\section{Bell's theorem}

In 1964, John S.~Bell published a bombshell paper showing that the
predictions of quantum theory are \textit{inherently inconsistent}
with hidden variable theories.  The amazing thing about this result,
which is known as \textbf{Bell's theorem}, is that it requires no
knowledge about the details of the hidden variable theory, except that
it is deterministic and local.  In this section, we study a simplified
version of Bell's theorem due to D.~Mermin (Mermin 1981).

We again consider spin-1/2 particle pairs, with particle $A$ on Earth
and particle $B$ in the Betelgeuse star system.  The experimentalists
at the two locations can measure the spin along three distinct choices
of spin axis, whose spin observables are denoted by $S_1$, $S_2$,
$S_3$.  We will not specify the actual directions of these spin axes
until later in the proof.  (Note, however, that they need not
correspond to orthogonal spatial directions.)

In a hidden variables theory, each particle must have a definite value
of each spin observable.  For example, particle $A$ might have spin
values $S_1 = +\hbar/2, \, S_2 = +\hbar/2, \, S_3 = -\hbar/2$.  For
conciseness, we denote this by $[++-]$.

We now imagine repeatedly preparing two-particle systems, sending the
particles to Earth and Betelgeuse, and letting the two
experimentalists measure the spin of each particle.  Within the
framework of quantum mechanics, the system is prepared each time in
the two-particle state
$$|\psi\rangle = \frac{1}{\sqrt{2}} \Big(|\!+\!z\rangle\otimes|\!-\!z\rangle \,-\, |\!-\!z\rangle\otimes|\!+\!z\rangle\Big).$$
Moreover, in each instance of the experiment, each experimentalist
independently and randomly chooses one of the three possible axes for
measuring spin.  (It doesn't matter which experimentalist performs the
measurement first.)  The procedure is repeated many times, and both
experimentalists' choices of spin axis are recorded, along with their
measurement results.

\textcolor{red}{[Fig.]}

At the end of the run of experiments, the records are examined.  The
results are found to be consistent with the predictions of quantum
theory.  In particular, whenever the experimentalists happen to
choose the same spin axis (e.g., both measuring $S_1$), they always
find opposite spins.

We are interested in the hypothesis that quantum theory is a
statistical description of an underlying hidden variables theory.  To
be consistent with the predictions of quantum theory, the hidden spin
states of the two particles must have opposite values along each
direction.  This means that there are $8$ distinct possibilities,
which we can denote as
$$\begin{aligned}{[}{+++};{---}], \;\;\; [{++-};{--+}], \;\;\; [{+-+};{-+-}], \;\;\; [{+--};{-++}],\\ [{-++};{+--}], \;\;\; [{-+-};{+-+}], \;\;\; [{--+};{++-}], \;\;\; [{---};{+++}].\end{aligned}$$
For instance, $[{++-};{--+}]$ indicates that for particle $A$, $S_1 =
S_2 = +\hbar/2$ and $S_3 = -\hbar/2$, while particle $B$ has the
opposite spin values, $S_1 = S_2 = -\hbar/2$ and $S_3 = +\hbar/2$.
However, we don't know anything about the relative probabilities of
these 8 cases.

Let's now focus on the subset of experiments in which the two
experimentalists happened to choose \textit{different} spin axes
(e.g., Earth chose $S_1$ and Betelgeuse chose $S_2$).  Within this
subset, what is the probability \textit{for the two measurement
  results to have opposite signs} (i.e., one $+$ and one $-$)?  To
answer this question, we first look at the following 6 cases:
$$\begin{aligned}{[}{++-};{--+}], \;\;\; [{+-+};{-+-}], \;\;\; [{+--};{-++}],\\ [{-++};{+--}], \;\;\; [{-+-};{+-+}], \;\;\; [{--+};{++-}].\end{aligned}$$
These are the cases which do not have all $+$ or all $-$ for each
particle.  Consider one of these, say ${[}{++-};{--+}]$.  The two
experimentalists picked their measurement axes at random each time,
and amongst the experiments where they picked different axes, there
are only two ways for the measurement results to have opposite signs:
$(S_1,S_2)$ or $(S_2,S_1)$.  There are four ways to get the same sign:
$(S_1,S_3)$, $(S_2,S_3)$, $(S_3,S_1)$ and $(S_3, S_2)$.  Thus,
\textit{for this particular set of spin values}, the probability for
measurement results with opposite signs is 1/3.  If we go through all
6 of the cases listed above, we find the same conclusion: the
probability for opposite signs is 1/3.

Now look at the remaining 2 cases:
$${[}{+++};{---}], \;\;\; [{---};{+++}].$$
For these, the experimentalists will find results that have opposite
signs with probability 1.  Putting these results together, we arrive
at the following statement:

\textit{Given that the two experimentalists choose different spin
  axes, the probability that their results have opposite signs is $P
  \ge 1/3$.}

This is called \textbf{Bell's inequality}.  If there is ever a
situation where quantum theory predicts a probability $P < 1/3$ (i.e.,
a violation of Bell's inequality), we will be forced to conclude that
quantum theory is inherently inconsistent with local deterministic
hidden variables.  Note that we don't need to know anything about the
``inner workings'' of the hidden variables theory; in the above
derivation of Bell's inequality, we did not assume anything about the
relative probabilities for the various spin values.

To complete the proof, we need to find a set of $S_1$, $S_2$, and
$S_3$ such that the predictions of quantum mechanics violate Bell's
inequality.  One simple choice is to align $S_1$ with the $\hat{z}$
axis, and align $S_2$ and $S_3$ along the $\hat{x}$-$\hat{z}$ plane at
$120^\circ$ ($2\pi/3$ radians) from $S_1$, as shown in the figure:

\textcolor{red}{[Fig.]}

Thus, the spin operators can be written in the $\hat{S}_z$ eigenbasis as
$$\begin{aligned}\hat{S}_1 &= \frac{\hbar}{2} \, \sigma_3 \\ \hat{S}_2 &= \frac{\hbar}{2} \, \left[\cos(2\pi/3) \sigma_3 + \sin(2\pi/3)\sigma_1\right]  \\   \hat{S}_3 &= \frac{\hbar}{2} \, \left[\cos(2\pi/3) \sigma_3 - \sin(2\pi/3)\sigma_1\right].\end{aligned}$$

Suppose the Earth experimentalist (measuring $A$) chooses $S_1$, and
obtains $+\hbar/2$.  The state of particle $A$ collapses to
$|\!+\!z\rangle$, and the state of particle $B$ collapses to
$|\!-\!z\rangle$.  The Betelgeuse experimentalist is assumed to
choose a different spin axis.  If the choice is $S_2$, the expectation
value of the measurement is
$$\begin{aligned}\langle\, - z \, | \, S_2 \,|-\!z\,\rangle &= \frac{\hbar}{2} \Big[\cos(2\pi/3) \langle\,- z\,|\sigma_3| - \!z\,\rangle + \sin(2\pi/3)\langle\,- z\,|\sigma_1|-\!z\,\rangle\Big]\\ &= \frac{\hbar}{2} \cdot \frac{1}{2} \end{aligned}$$
If $P_+$ and $P_-$ respectively denote the probability of measuring
$+\hbar/2$ and $-\hbar/2$ in this measurement, the above equation
implies that $P_+ - P_- = + 1/2$.  Moreover, $P_+ + P_- = 1$ by
probability conservation.  It then follows that the probability of
obtaining a negative value (the opposite sign from the Earth
measurement) is $P_- = 1/4$.  All the other possible scenarios are
worked out in a similar way.  The overall conclusion is that if the
two experimentalists choose different measurement axis, they obtain
results of opposite signs with probability $1/4$.  This is smaller
than $1/3$, and hence violates Bell's inequality.

Finally, we must consult Nature by performing an experimental test.
Is it possible to observe, in an actual experiment, probabilities that
violate Bell's inequality?  In the decades following Bell's 1964
paper, many experiments were performed to answer this question.  These
experiments are all substantially more complicated than the simple
two-particle spin-$1/2$ model that we've studied, and they are subject
to various uncertainties and ``loopholes'' that are beyond the scope
of our discussion.  Overall, the experimental consensus is a clear
\textit{yes}: Nature really does behave as quantum mechanics predict,
in a manner that cannot be replicated using deterministic local hidden
variables!  For a review of the experimental evidence, see Aspect
(1999).

\section{Entanglement entropy}

In previous sections, we said that a multi-particle system is
``entangled'' if the individual particles do not have definite quantum
states.  It would be nice to make this statement more precise, by
formulating an expression for the level of entanglement in any given
system.  In fact, physicists have come up with several ways of
quantifying entanglement; in this section, we will describe the most
commonly-used one, called \textbf{entanglement entropy}, which is
closely related to the concept of entropy in thermodynamics,
statistical mechanics, and information theory.

Consider a quantum system with a $d$-dimensional Hilbert space
$\mathscr{H}$.  Given an arbitrary state $|\psi\rangle \in
\mathscr{H}$, we can define an operator
$$\hat{\rho}(\psi) = |\psi\rangle\, \langle\psi|.$$
This is simply the projection operator associated with $|\psi\rangle$,
but in this context we call it a \textbf{density matrix}.  (In the
language of linear algebra, $\hat{\rho}(\psi)$ is formed by taking the
``matrix outer product'' of the vector $|\psi\rangle$ with its
Hermitian conjugate.)  We use the notation $\hat{\rho}(\psi)$ to
emphasize that the density matrix is dependent on the chosen state
$|\psi\rangle$.  Evidently, it is a Hermitian operator; one eigenvalue
is 1 (with eigenvector $|\psi\rangle$), and the other $d-1$
eigenvalues are all $0$ (with eigenvectors given by $d-1$ vectors
spanning the subspace orthogonal to $|\psi\rangle$).

The expectation values of the density matrix have a special meaning.
Suppose the system is in state $|\psi\rangle$, and we perform a
measurement corresponding to a Hermitian operator $\hat{Q}$, which has
eigenvalues $\{q_1,q_2,\dots\}$ and eigenvectors
$\{|q_1\rangle,|q_2\rangle,\dots\}$.  Then the probability of
measuring $q_j$ is
$$P_j \;=\; |\langle q_j| \psi\rangle|^2 \;=\; \langle q_j |\psi\rangle \langle \psi|q_j\rangle \;=\; \langle q_j |\, \hat{\rho}(\psi)\, |q_j \rangle.$$
In other words, the expectation values of $\hat{\rho}(\psi)$
correspond to the probabilities for various outcomes of measurements
performed on the state $|\psi\rangle$.

Now suppose $\mathscr{H} = \mathscr{H}_A \otimes \mathscr{H}_B$, where
$\mathscr{H}_A$ and $\mathscr{H}_B$ are the Hilbert spaces of two
subsystems.  Starting from a density matrix of the combined system,
$\hat{\rho}(\psi)$, we can define a \textbf{``reduced'' density
  matrix} for subsystem A:
$$\hat{\rho}_A(\psi) = \mathrm{Tr}_B \,\big[\,\hat{\rho}(\psi)\,\big].$$
Here, $\mathrm{Tr}_B[\cdots]$ refers to a \textbf{partial trace},
tracing over the Hilbert space of subsystem B.  What's left after
this partial trace is an operator acting solely on the Hilbert space
$\mathscr{H}_A$.

To understand the meaning of $\hat{\rho}_A(\psi)$, let us work in
an explicit basis.  Let $\{|\mu_1\rangle, |\mu_2\rangle,\cdots\}$ be a
basis for $\mathscr{H}_A$, and $\{|\nu_1\rangle,
|\nu_2\rangle,\cdots\}$ a basis for $\mathscr{H}_B$.  A given state of
the combined system can be written in the tensor product basis:
$$|\psi\rangle = \sum_{ij} \psi_{ij} \, |\mu_i\rangle \otimes |\nu_j\rangle, \;\;\; \mathrm{where}\;\; \psi_{ij} \in \mathbb{C}.$$
The density matrix of the combined system is
$$\hat{\rho}(\psi) = |\psi\rangle \langle\psi| = \sum_{ii'jj'} \Big( |\mu_i\rangle \otimes |\nu_j\rangle\Big) \psi_{ij} \psi_{i'j'}^* \, \Big(\langle \mu_{i'}| \otimes \langle \nu_{j'}|\Big).$$
The density matrix for subsystem A is then obtained by tracing over
the states of subsystem B:
$$\begin{aligned}\hat{\rho}_A(\psi) &= \sum_j \langle \nu_j | \hat{\rho}(\psi) | \nu_j \rangle \\ &= \sum_{ii'} |\mu_i\rangle \,\rho_{ii'}^A \, \langle \mu_{i'}| \;\;\;\mathrm{where} \;\; \rho_{ii'}^A \equiv \sum_j \psi_{ij} \psi_{i'j}^*.\end{aligned}$$
Note that $\hat{\rho}_A(\psi)$ is clearly Hermitian.

The density matrix for subsystem A contains information about the
outcome probabilities for \textit{partial} measurements performed on
subsystem A.  Suppose the overall system is in state $|\psi\rangle$,
and we perform a measurement on subsystem A corresponding to a
Hermitian operator $\hat{\mu}$, which has eigenvectors
$\{|\mu_1\rangle,|\mu_2\rangle,\dots\}$.  Then the probability of
measuring $\mu_i$ is
$$\langle \mu_i | \hat{\rho}_A |\mu_i \rangle \,=\, \rho_{ii}^A \,=\, \sum_{j} |\psi_{ij}|^2.$$
(Note that this is consistent with the discussion of partial
measurements in Section~\ref{sec:partialmeasurements}.)  In other
words, the diagonal matrix elements of $\hat{\rho}_A$, in any basis,
specify the outcome probabilities for measurements corresponding to
that basis.  Conservation of probability implies that $\sum_i
\rho^A_{ii} = 1$, and hence that
$$\mathrm{Tr}\big[\rho_A\big] = 1.$$

In entangled systems, the density matrix essentially takes over the
theoretical role of the quantum state.  In a single-particle or
unentangled system, we can define the quantum state and use it to
calculate the probabilities of various measurement outcomes.  However,
in a system that is entangled with other subsystems of a larger
system, there is no definite quantum state.  Instead, probabilities
are calculated from the density matrix.

Given a density matrix, we can define a quantity called the
\textbf{entanglement entropy}:
$$S = - k_B \mathrm{Tr} \Big\{ \hat{\rho}\, \ln\!\big[\hat{\rho}\big] \Big\}.$$
Here, $S$ refers to the entropy of the system or subsystem described
by the density matrix $\hat{\rho}$; for instance, if we use the
density matrix $\hat{\rho}_A$ in the above formula, then the formula
gives the entropy of subsystem A.  We use $\ln[\cdots]$ to denote the
logarithm of an operator, which is basically the inverse of the
exponential: $\ln(\hat{A}) = \hat{B} \Rightarrow \exp(\hat{B}) =
\hat{A}$.  The prefactor $k_B$ is Boltzmann's constant, and is present
to ensure that $S$ has the same units as thermodynamic entropy.

The definition of the entanglement entropy is based on an analogy with
the entropy concept from thermodynamics, statistical mechanics and
information theory.  In those classical contexts, entropy is a
quantitative measure of uncertainty (i.e, lack of information)
about a system's underlying microscopic state, or ``microstate''.
Suppose a system has $W$ possible microstates that occur with
probabilities $\{p_1, p_2, \dots, p_W\}$, satisfying $\sum_i p_i = 1$.
Then we define the classical entropy as
$$S_{\mathrm{classical}} = - k_B \sum_{i=1}^W p_i \ln(p_i).$$
If the system is definitely in one microstate $k$ (i.e., $p_i =
\delta_{ik}$), a situation of complete certainty, the above formula
gives $S_{\mathrm{classical}} = 0$.  (Note that $x \ln(x)\rightarrow
0$ as $x\rightarrow 0$).  On the other hand, in a situation of
complete uncertainty where all microstates are equally probable ($p_i
= 1/W$ for all $i$), the formula gives $S_{\mathrm{classical}} = k_B
\ln(W)$, which is the entropy of a microcanonical ensemble in
statistical mechanics.  For any other distribution of probabilities,
it can be shown that the entropy lies between these two extremes: $0
\le S_{\mathrm{classical}} \le k_B\ln(W)$.

Entanglement entropy quantifies the uncertainty arising from an
entangled system's lack of a definite quantum state.  But we need to
be careful when extending classical notions of probability to quantum
systems.  As described above, when performing a measurement whose
possible outcomes are $\{\mu_1, \mu_2, \dots\}$, the probability of
getting $\mu_i$ is $p_i = \langle \mu_i | \hat{\rho}|\mu_i\rangle$.
However, it is problematic to directly substitute these probabilities
$\{p_i\}$ into the classical entropy formula, since they are
basis-dependent (i.e., the set of probabilities is dependent on the
choice of measurement).  The above definition for the entanglement
entropy, $S = - k_B \mathrm{Tr} \big\{ \hat{\rho}\, \ln[\hat{\rho}]
\big\}$, bypasses this problem by using the trace, which is
basis-independent.  This also means that if we look specifically at
the eigenbasis for $\hat{\rho}$, denoted by $\{|\mu'_i\rangle\}$, then
$$\begin{aligned}S &= -k_B \sum_i \langle \mu'_i | \hat{\rho}\ln(\hat{\rho}) | \mu'_i\rangle  \\ &= - k_B \sum_i p_i' \ln(p_i') \;\;\;\mathrm{where}\;\;\hat{\rho}\,|\mu'_i\rangle = p_i' |\mu'_i\rangle\end{aligned}$$
In this specific basis, where the eigenvalues $p_i =
\langle\mu'_i|\hat{\rho}|\mu'_i\rangle$ are also the measurement
probabilities, the definition of the entanglement entropy is
consistent with the definition of the classical entropy.

For an unentangled system, we can write $\hat{\rho} =
|\psi\rangle\langle\psi|$ for some state $|\psi\rangle$, so
$|\psi\rangle$ is itself an eigenstate of $\hat{\rho}$ with eigenvalue
1; hence, $S = 0$.  Conversely, if we find that a system has $S \ne
0$, that implies that it is entangled, and lacks a definite quantum
state.  A system is \textbf{maximally entangled} if the density matrix
eigenvalues (measurement probabilities) are all equal, $p_i' = 1/d$,
where $d$ is the dimension of the Hilbert space, so that $S = k_B
\ln(d)$.  In general, the entanglement entropy satisfies $0 \le S \le
k_B\ln(d)$.

Let us apply the entanglement entropy formula to the system of two
spin-$1/2$ systems and the state
$$|\psi\rangle = \frac{1}{\sqrt{2}} \Big(|\!+\!z\rangle\otimes|\!-\!z\rangle \,-\, |\!-\!z\rangle\otimes|\!+\!z\rangle\Big).$$
The density matrix for the two-particle system is
$$\hat{\rho}(\psi) = \frac{1}{2} \Big(|\!+\!z\rangle\otimes|\!-\!z\rangle \,-\, |\!-\!z\rangle\otimes|\!+\!z\rangle\Big) \Big(\langle+z|\otimes\langle-z| \,-\, \langle-z|\otimes\langle+z|\Big).$$
Tracing over the B subsystem gives the reduced density matrix:
$$\hat{\rho}_A(\psi) = \frac{1}{2} \Big(|\!+\!z\rangle \langle+z| \,+\, |\!-\!z\rangle \langle-z|\Big).$$
This can be expressed as a matrix in the
$\{|\!+z\rangle,|\!-z\rangle\}$ basis:
$$\hat{\rho}_A(\psi) = \begin{pmatrix}\frac{1}{2} & 0 \\ 0 & \frac{1}{2}\end{pmatrix}.$$
Note that this satisfies $\mathrm{Tr}(\hat\rho_A) = 1$, as expected.
Next, we can use $\hat{\rho}_A$ to compute the entropy of subsystem A:
$$S_A = -k_B\mathrm{Tr}\left\{\hat{\rho}_A\ln(\rho_A)\right\} = -k_B\mathrm{Tr}\begin{pmatrix}\frac{1}{2}\ln\left(\frac{1}{2}\right) & 0 \\ 0 & \frac{1}{2}\ln\left(\frac{1}{2}\right)\end{pmatrix} = k_B\ln(2).$$
This is the maximum possible entanglement entropy for a system with a
2D Hilbert space.  Hence, subsystems A and B are maximally entangled.

\section{The Many-Worlds Interpretation}

We conclude this chapter by discussing a logical but controversial set
of implications that arise from the phenomenon of quantum
entanglement: the \textbf{Many-Worlds Interpretation} of quantum
theory.

In describing the phenomenon of state collapse, we have relied heavily
on the \textbf{measurement postulate} of quantum mechanics, which
states: ``If a system has state $|\psi\rangle$, and we perform a
measurement $\hat{Q}$, the probability of a result $q$ (which is an
eigenvalue of $\hat{Q}$ with eigenvector $|q\rangle$) is $P_q =
|\langle q | \psi\rangle|^2$.  After obtaining this measurement
result, the quantum state collapses to $|q\rangle$.''

The measurement postulate is part of the \textbf{Copenhagen
  Interpretation} of quantum mechanics.  This is how quantum mechanics
is typically taught, and how physicists think about the theory when
doing practical, everyday calculations.  However, it has two
unsatisfactory features.

First, the measurement postulate seems to somehow stand apart from the
other postulates of quantum mechanics.  It is the only place where
randomness, or ``indeterminism'', creeps into quantum theory.  The
other postulates do not refer to probabilities; the Schr\"odinger
equation
$$i\hbar\frac{\partial}{\partial t}|\psi(t)\rangle = \hat{H}(t) |\psi(t)\rangle$$
is completely deterministic, in the sense that if I tell you that the
state vector at time $t_0$ is $|\psi(t_0)\rangle$, then knowing the
Hamiltonian $\hat{H}(t)$ you can find $|\psi(t)\rangle$ for all $t$ by
solving the Schr\"odinger equation.  This time-evolution consists of a
smooth, non-random rotation of the state vector within its Hilbert
space.  A measurement, however, has a completely different effect: it
causes the state vector to instantaneously jump to a randomly-selected
value.  It is strange for a theory to contain two fundamentally
different ways for a state to change.

The second noteworthy feature of the measurement postulate is that it
is silent on what exactly constitutes a ``measurement''.  Does a
measurement require a conscious observer?  Surely not: in the words of
Einstein, could we really believe that the Moon exists only when we
look at it?  Given a device that interacts with a particle, how do we
figure out whether the interaction amounts to a measurement, or
whether it affects the particle via the Schr\"odinger equation?

The Many-Worlds Interpretation seeks to resolve these problems by
positing that \textit{the measurement postulate is \underline{not} a
  fundamental postulate of quantum mechanics}.  Rather, what we call
``measurement'' (including state collapse and the apparent randomness
of measurement results) is an emergent phenomenon, which can be
derived from the behavior of complex many-particle quantum systems
obeying the Schr\"odinger equation.

The key idea is to model a measurement process by applying the
Schr\"odinger equation to a multi-particle quantum system, one that
contains both the particle being measured \textit{and} the measurement
apparatus itself.  Suppose, for example, that we have a spin-$1/2$
particle, and an apparatus designed to measure $S_z$.  Let
$\mathscr{H}_p$ denote the 2D Hilbert space for the particle's
spin-$1/2$ degree of freedom.  Let $\mathscr{H}_A$ denote the Hilbert
space of the apparatus, which is \textit{much} larger and more
complicated than $\mathscr{H}_p$ (since it has to describe the
$10^{23}$ or more atoms in the apparatus).  The Hilbert space of the
combined system can be written as
$$\mathscr{H} = \mathscr{H}_p \otimes \mathscr{H}_A.$$
The system is prepared in an initial state of the form
$$|\psi(0)\rangle = |\!+x\rangle \otimes |\textbf{START}\rangle,$$
where $|\!+x\rangle \in \mathscr{H}_p$ is an initial state of the
particle that does not have a definite value of $S_z$, while
$|\textbf{START}\rangle \in \mathscr{H}_A$ denotes the initial state
of the apparatus.

The combined system then undergoes time evolution via the
Schr\"odinger equation, with an extremely complicated Hamiltonian that
describes how the apparatus acts upon the particle to perform an $S_z$
measurement.  After some time $T$ has passed, the state evolves to
$$|\psi(T)\rangle \;=\; |\!+z \rangle \otimes |\textbf{RESULT}\uparrow\rangle \;+\; |\!-z\rangle\otimes |\textbf{RESULT}\downarrow\rangle, $$
where $|\textbf{RESULT}\uparrow\rangle, \,
|\textbf{RESULT}\downarrow\rangle \in \mathscr{H}_A$ are complicated
multi-particle state vectors of the apparatus.  For example, the
apparatus might have a centimeter-size pointer (of $\sim 10^{23}$
atoms) that indicates the measurement result;
$|\textbf{RESULT}\uparrow\rangle$ describes the apparatus with the
pointer pointing up (indicating a $+z$ result), while
$|\textbf{RESULT}\downarrow\rangle$ describes the apparatus with the
pointer pointing down (indicating a $-z$ result).

In the above expression for $|\psi(T)\rangle$, the two terms can be
interpreted as two ``alternative histories'' of the system, or two
different ``worlds''.  In one world, the pointer on the apparatus
indicates a $+z$ measurement result, and the quantum state of the
particle has ``collapsed'' to $|\!+z\rangle$.  In the other world, the
pointer indicates a $-z$ result, and the particle state has
``collapsed'' to $|\!-z\rangle$.  Moreover, with a bit more thought we
can work out that the relative weights of the two terms correspond to
the probabilities of obtaining each measurement result.  In other
words, the effects of the measurement postulate (i.e., random
measurement results and the phenomenon of state collapse) arise from
restricting our attentions to a single world.

\textcolor{red}{Explicit model?}

This description can be generalized to the universe as a whole.  In
the viewpoint of the Many-Worlds Interpretation, the entire universe
can be described by a mind-bogglingly complicated quantum state,
evolving deterministically according to the Schr\"odinger equation.
This evolution involves repeated ``branchings'' of the universal
quantum state, each producing more and more worlds.  The classical
world that we appear to inhabit is just one amongst an
incomprehensibly vast multitude of worlds.  It is up to you to decide
whether this conception of reality makes sense.  It is essentially a
matter of preference, because the Copenhangen Interpretation and the
Many-Worlds Interpretation have identical physical consequences (which
is why they are referred to as different ``interpretations'' of
quantum mechanics, rather than different theories).




\section{Exercises}

%% \begin{enumerate}
%% \end{enumerate}


\section{Further Reading}

\begin{itemize}
\item A.~Einstein, B.~Podolsky, and N.~Rosen,
  \textit{Can Quantum-Mechanical Description of Physical Reality Be
    Considered Complete?}, Physical Review \textbf{47}, 777 (1935).

\item J.~S.~Bell, \textit{On the Einstein-Podolsky-Rosen paradox},
  Physics \textbf{1}, 195 (1964).
  
\item N.~D.~Mermin, \textit{Bringing home the atomic world: Quantum
  mysteries for anybody}, American Journal of Physics \textbf{49}, 940
  (1981).

\item A.~Aspect, \textit{Bell's inequality test: more ideal than ever},
  Nature (News and Views) \textbf{398}, 189 (1999).
\end{itemize}

\end{document}


%% For decades after the discovery of quantum mechanics, the quantum
%% double-slit experiment was just a ``thought experiment'', meant to
%% illustrate the features of quantum mechanics that had been uncovered
%% by other, more complicated experiments.  Nowadays, the most convenient
%% way to do the experiment is with light, using single-photon sources
%% and single-photon detectors.  Quantum interference has also been
%% demonstrated experimentally using electrons, neutrons, and even
%% large-scale particles such as buckyballs.
