\documentclass[pra,12pt]{revtex4}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{color}
\usepackage{mathrsfs}
\usepackage[pdfborder={0 0 0},colorlinks=true,linkcolor=blue,urlcolor=blue]{hyperref}

\def\ket#1{\left|#1\right\rangle}
\def\bra#1{\left\langle#1\right|}
\def\braket#1{\left\langle#1\right\rangle}

\setlength{\parindent}{0pt}

\renewcommand{\baselinestretch}{1.0}
\setlength{\parskip}{0.07in}

\begin{document}

\begin{center}
{\large \textbf{Appendix D: Numerical Tensor Products}}
\end{center}

As described in Chapter 2, multi-particle quantum states are described
mathematically using tensor products.  This appendix discusses how
tensor product states, and the operators acting upon them, are handled
in numerical linear algebra software like
\href{https://scipy.org/}{Scientific Python}, GNU Octave, or Matlab.

Numerical linear algebra is always performed in an explicit basis, so
that states are represented by arrays of vector components, and linear
operators are represented by matrices.  Tensor products can be
represented by an operation known as the ``Kronecker product'', which
is implemented by the
\href{https://docs.scipy.org/doc/numpy-1.13.0/reference/generated/numpy.kron.html}{\texttt{numpy.kron}}
function in Scientific Python, and the \texttt{kron} function in GNU
Octave and Matlab.

The Kronecker product operation takes two inputs, $A$ and $B$, and
returns a new array consisting of blocks of $B$ scaled by the
components of $A$.  What this means is best understood in terms of the
two most common cases: (i) products of vectors, and (ii) products of
matrices.

\section{Product of two vectors}

First, suppose that $A$ and $B$ are two vectors (i.e.,
one-dimensional arrays arrays), of length $M$ and $N$ respectively;
let their components be $(A_0, A_1, \dots, A_{M-1})$ and $(B_0, B_1,
\dots, B_{N-1})$.  (Following the conventions of Scientific Python, we
will not distinguish between row and column vectors, and components
are indexed starting from 0.)  Then
$$\texttt{kron}(A, B) = \Big(A_0B_0,\,\dots,\, A_0 B_{N-1},\, A_1 B_0,\, \dots,\, A_1 B_{N-1},\, \dots,\, A_{M-1}B_{N-1}\Big).$$
This vector has length $MN$, and its first $N$ components are $A_0
(B_0, \dots,B_{N-1})$, the next $N$ components are $A_1 (B_0, \dots,
B_{N-1})$, etc.  We can describe this compactly in index notation:
$$\big[\, \texttt{kron}(A, B) \,\big]_{\mu} = A_m \, B_n \;\;\;\mathrm{where}\;\;\;\mu = mN+n.$$
Note that the index $\mu$ sweeps through the values $\mu =
0,1,\dots,MN-1$, without duplication, as we take $m = 0,\dots,M-1$ and
$n = 0,\dots,N-1$.

This operation can be used to describe the tensor product of two
quantum state vectors.  Suppose we have a state vector $|A\rangle$ in
an $M$-dimensional Hilbert space $\mathscr{H}_A$, and a state vector
$|B\rangle$ in an $N$-dimensional Hilbert space $\mathscr{H}_B$.  Let
the two Hilbert spaces have bases $\{|m\rangle\}$ and $\{|n\rangle\}$
respectively.  The state vectors can be written using the basis
decompositions
$$|A\rangle = \sum_{m=0}^{M-1} A_m |m\rangle, \quad |B\rangle = \sum_{n=0}^{N-1} B_n |n\rangle.$$
The tensor product $|A\rangle\otimes|B\rangle$ is a vector in the
space $\mathscr{H}_A\otimes \mathscr{H}_B$, which is $MN$-dimensional.
A natural basis for this product space is
$$\Big\{|\mu\rangle \equiv |m\rangle\otimes |n\rangle\Big\} \;\;\;\mathrm{where} \;\begin{cases}\mu\!\!\!\! &= mN+n \\ m \!\!\!\!&= 0,1,\dots,M-1 \\ n \!\!\!\!&= 0,1, \dots, N-1.\end{cases}$$
Then the components of $|A\rangle\otimes|B\rangle$ in this basis are
given by the Kronecker product:
$$|A\rangle\otimes|B\rangle = \sum_{\mu=0}^{MN-1} C_\mu \, |\mu\rangle, \;\;\; \mathrm{where}\;\;\; C_\mu\, = \big[\,\texttt{kron}(A,B)\,\big]_\mu.$$

\section{Product of two matrices}

Consider the case where $A$ is an $M\times M$ matrix and $B$ is an
$N\times N$ matrix:
$$A = \begin{bmatrix}A_{00} & \cdots & A_{0,M-1} \\ \vdots & \ddots & \vdots \\ A_{M-1,0} & \cdots & A_{M-1,M-1} \end{bmatrix}, \;\; B = \begin{bmatrix}B_{00} & \cdots & B_{0,N-1} \\ \vdots & \ddots & \vdots \\  B_{N-1,0} & \cdots & B_{N-1,N-1} \end{bmatrix}.$$
In this case, the Kronecker product of $A$ and $B$ is an $MN\times MN$
matrix of the form
$$\texttt{kron}(A,B) = \begin{bmatrix} A_{00}B & \cdots & A_{0,M-1}B \\ \vdots & \ddots & \vdots \\ A_{M-1,0}B & \cdots & A_{M-1,M-1}B\end{bmatrix}.$$
In component terms,
$$\big[\,\texttt{kron}(A,B)\,\big]_{\mu\mu'} = A_{mm'} B_{nn'}\;\;\;\mathrm{where}\;\;\;\mu = mN+n, \; \mu' = m'N+n'.$$
The index $\mu$ sweeps through the values $\mu = 0,1,\dots,MN-1$,
without duplication, as we take $m = 0,\dots,M-1$ and $n =
0,\dots,N-1$; and likewise for the primed indices.

The Kronecker product of two matrices can be used to describe the
tensor product of two quantum operators.  Let $\hat{A}$ be a quantum
operator acting on the $M$-dimensional Hilbert space $\mathscr{H}_A$,
and $\hat{B}$ a quantum operator acting on the $N$-dimensional Hilbert
space $\mathscr{H}_B$.  As before, let the bases for the two spaces be
$\{|m\rangle\}$ and $\{|n\rangle\}$ respectively, so that
$$\hat{A} = \sum_{m,m'=0}^{M-1}  |m\rangle A_{mm'} \langle m'|, \quad \hat{B} = \sum_{n,n'=0}^{N-1} |n\rangle B_{nn'}\langle n'|.$$
The tensor product $\hat{A}\otimes\hat{B}$ acts on the
$MN$-dimensional Hilbert space $\mathscr{H}_A\otimes \mathscr{H}_B$.
Again, we adopt the basis $\big\{|\mu\rangle \equiv |m\rangle\otimes
|n\rangle \big\}$ where $\mu = mN+n$; then
$$\hat{A}\otimes\hat{B} = \sum_{\mu,\mu'=0}^{MN-1} |\mu\rangle C_{\mu\mu'} \langle\mu'| \;\;\; \mathrm{where}\;\;\; C_{\mu\mu'}\, = \big[\,\texttt{kron}(A,B)\,\big]_{\mu\mu'}.$$

Thus, we see that the Kronecker product takes pairs of vectors, or
pairs of linear operators, and returns the tensor product components
corresponding to a particular basis choice for the product space.
Note that the order of inputs matters: $\texttt{kron}(A,B)$ is not the
same as $\texttt{kron}(B,A)$, since the inputs determine the ordering
of the basis vectors in the product space (in the former case, $\mu =
mN+n$ rather than $\mu = nM+m$).  Therefore, we must stick to one
choice throughout a program (e.g., vectors and matrices for Hilbert
space $\mathscr{H}_A$ always go in the first slot, and those for
Hilbert space $\mathscr{H}_B$ in the second slot).

\section{Projections and partial traces}


\section{Example: coupled spin-$\frac{1}{2}$ systems}

\end{document}
